---
title: "Developer guide"
description: >
  Status, running from a clone, tests, and the project layout.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developer guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
```

This page collects the contributor-facing details. For the design rationale and
decision log, see
[`ARCHITECTURE_REVIEW.md`](https://github.com/stanstrup/xcmsVisGUI/blob/main/ARCHITECTURE_REVIEW.md)
in the repository.

## Status

Working: async file ingestion; settings (ColorBrewer/viridis palette, retention-time
unit, default EIC tolerance, export defaults — persisted across restarts); global
filters (rt / *m/z* / MS level / polarity / intensity / repeatable spectrum-id
rules); per-view profile-mode peak picking (`Spectra::pickPeaks`, with S/N /
half-window / m/z-refinement); and all raw-data plot views — TIC/BPC, multi-EIC
(with intensity scaling), click-to-spectrum (+ scan-list browser), 2D MS map, 3D
points/surface, and DDA precursor ions. The Spectrum view also does single-spectrum
adduct / isotope / in-source-fragment annotation (manual anchor, findMAIN
auto-suggest, difference network, and formula-based fine isotope patterns via
Rdisop + enviPat).

Scope is **raw visualisation** — profile scans are centroided on the fly for
display/annotation, but there is no cross-sample peak picking / feature grouping /
retention-time alignment (deferred; see the architecture doc).

Extraction results are cached to disk (qs2), so re-opening the app with the same
files + filter is instant. Figures export as png/svg/pdf, or as the raw ggplot
object (`.rds`) for later tweaking in R.

## Run from a clone

This is an R package; the app is the exported `run_app()`. From a fresh clone,
install the dependencies once into your normal R library, then launch:

```{r, eval = FALSE, echo = TRUE}
# install the dependencies (Imports + Suggests + the commonMZ GitHub remote):
# install.packages(c("remotes", "BiocManager"))
# options(repos = BiocManager::repositories())   # so the Bioconductor deps resolve
# remotes::install_deps(dependencies = TRUE)

# then launch from source (load_all live-reloads your edits):
# Rscript run.R           # = pkgload::load_all() + run_app(launch.browser = TRUE)
```

`BiocManager` is needed so the Bioconductor dependencies (Spectra, xcms, mzR, …)
resolve; `remotes` follows the `Remotes:` field to install `commonMZ` from GitHub.
There's no renv to set up — `DESCRIPTION` is the single source of truth for
dependencies (the same way users install the package, and the same way CI
provisions via `r-lib/actions/setup-r-dependencies`).

## Tests

```{r, eval = FALSE, echo = TRUE}
# testthat::test_local(".")   # or: R CMD check
```

The suite covers the pure helpers and the key invariant that `apply_filters`
(MsExperiment) and `apply_filters_spectra` (Spectra) select the same spectra.
Real-data tests use the `msdata` / `faahKO` Bioconductor packages and skip if
absent. CI runs `R CMD check` on push/PR (`.github/workflows/R-CMD-check.yaml`).

## Regenerating the documentation screenshots

The article screenshots are captured headlessly with
[`chromote`](https://rstudio.github.io/chromote/) against a running app. Start the
app on port 7799, then run the capture script in a second process:

```{r, eval = FALSE, echo = TRUE}
# terminal 1 — serve the app on the port the script expects:
# Rscript -e "pkgload::load_all('.'); run_app(port = 7799, launch.browser = FALSE)"

# terminal 2 — drive it and write the PNGs:
# Rscript data-raw/capture-screenshots.R
```

They are saved at 1440x900 into `vignettes/articles/figures/`. Chrome only captures
raster images (PNG), so the UI screenshots are PNG; individual plots can instead be
exported as SVG from each view's **Save** button if a vector figure is wanted.

## Project layout

```
run.R                  # convenience launcher (load_all + run_app)
DESCRIPTION / NAMESPACE # package metadata; NAMESPACE is roxygen-generated
R/
  run_app.R            # app_ui() / app_server() + exported run_app()
  zzz.R                # .onLoad: register BiocParallel SerialParam (the perf fix)
  constants.R          # MS-file constants, palette names, rt-unit helpers
  daemons.R            # mirai daemon pool + per-run setup
  xcmsVisGUI-package.R # roxygen import declarations + globalVariables
  mod_ingest.R         # typed-path / choose.dir / fileInput + async mirai reader + file list
  mod_settings.R       # palette, rt unit, default tolerance, daemon count, export; persistence
  mod_filter.R         # global rt/mz/MS-level/polarity/intensity + spectrum-id rules
  mod_plot_tic_bpc.R   # TIC/BPC overlay, colour by group/sample, click->spectrum
  mod_plot_eic.R       # editable multi-m/z target table -> overlaid EICs
  mod_plot_spectrum.R  # spectrum at a clicked rt / picked scan + scan-list browser + annotation
  mod_plot_map.R       # 2D MS map + 3D points/surface (plotly-native)
  mod_plot_precursors.R# DDA precursor-ion map
  mod_export.R         # reusable png/svg/pdf/rds export modal
  fct_extract.R        # data extraction (summaries, chromatograms, peaks, spectra)
  fct_filters.R        # compose filter state into Spectra/xcms calls; profile peak-picking
  fct_annotate.R       # the single-spectrum annotation engine (pure, testable)
  fct_isotope.R        # fine isotope-pattern engine: Rdisop formulas + enviPat envelope
  fct_export.R         # ggsave-based export (+ rds = the ggplot object itself)
  fct_palettes.R       # ColorBrewer / viridis helpers
  fct_cache.R          # layered mem+disk (qs2) cache backing bindCache, persistent across restarts
  fct_settings_store.R # persist settings to the per-user config dir
  utils_reactive.R     # central reactive state (rv) + plotly/zoom + peak-picking UI helpers
tests/testthat/        # unit + real-data tests
data-raw/              # throwaway scripts, incl. capture-screenshots.R (chromote)
```
